Spike-and-Slab Generalized Additive Models and Fast Algorithms for High-Dimensional Data

### 2021-07-20 Manuscript/SS_GAM.tex
Correct AMA in-text citation

Using additive models to interpolate nonlinear effects has a long
history. Early work of additive models can date back to 1979 when
\citet{Cleveland1979} proposed local smoothing for bivariate analysis.
Since then, many mathematical and statistical models have been proposed,
see \citet{Wood2020}. With the recent advancement of computational
technology, many machine learning algorithms that are computationally
intensive start to gain much popularity in prognostic modeling and
functional selection, especially when complex signals, for example
nonlinear effects, are assumed. Such models generally have improved
prediction accuracy over classic statistical models. However, these
models have a fatal flaw, the lack of transparency and interpretability,
and hence are referred to as ``black box'' models. Given that the
primary objective of biomedical studies is to understand the etiology of
diseases, ``Black box'' models are less helpful as they can hardly
explain how the change of biomarkers is associated with the disease.

Semi-parametric regression models are a more appropriate alternative to
the ``black-box'' models, thanks to their balancing between model
flexibility and interpretability. Among those, generalized additive
models (GAMs), proposed in the seminal work of \citet{Hastie1987}, grew
to be one of the most utilized modeling tools. In a GAM, the response
variable, \(Y\), which is assumed to follow some exponential family
distribution \(EF(\mu, \phi)\)\footnote{TODO: explain \(\phi\)}, can be
expressed as the summation of smooth functions,
\(f_j, j = 1, \dots, p\), of a given \(p\)-dimensional vector of
covariates \(\boldsymbol{x}\), written as \[
E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}f_j(x_j)),
\] where \(g^{-1}\) is the inverse function of a smooth monotonic link
function. The smooth functions can take many forms and are estimated
using a pseudo-weighted version of the backfitting algorithm
\citep{Breiman1985}. Nevertheless, the classic GAMs cannot fulfill the
increasing analytic demands for high-dimensional data analysis in
biomedical studies.

Many applications of modern biomedical research, e.g.~sequencing data
analysis, electric health record data analysis, require special
treatment of high-dimensionality, commonly known as \(p >> n\) problem.
There is extensive literature on high-dimensional linear models via
penalized models or Bayesian hierarchical models, see
\citet{Mallick2013}. These models assume sparsity among the predictors,
i.e.~most predictors have small or no effects on the outcome. Further
development focuses on addressing the grouping structure among the
predictors, where AMs and GAMs can be considered special cases. Various
penalties are imposed on the basis function coefficients:
\citet{Ravikumar2009} extended the grouped lasso \citep{Yuan2006} to
additive models; \citet{Huang2010} further developed adaptive grouped
lasso for additive models; \citet{Wang2007} and \citet{Xue2009} applied
grouped SCAD penalty \citep{Fan2001} to the additive models. Recent
Bayesian hierarchical models are also used in the context of
high-dimensional additive models. Various group spike-and-slab priors
combining with computationally intensive Markov chain Monte Carlo (MCMC)
algorithms \citep{Xu2015, Yang2020} are proposed where the application
on AMs are by-products. \citet{Scheipl2012} proposed an ad hoc
spike-and-slab structure prior, normal-mixture-of-inverse gamma,
specifically for semi-parametric regression using MCMC. The
normal-mixture-of-inverse gamma prior is mixture of t-distribution with
an additional coefficients mixing vector, where the vector follows a
mixture of two Gaussian distribution. \citet{Bai2020} was the first to
apply group spike-and-slab lasso prior to AMs using a fast optimization
algorithm, and further generalized the framework to GAMs in
\citet{Bai2021}. Regardless of the nice statistical properties, a
practical and pivotal question, if non-linear effects exist, yet to be
addressed by these methods. Besides \citet{Schipel2012}, most proposed
methods adapt an `all-in-all-out' strategy, i.e.~either including or
excluding the variable completely, rendering no space for within group
selection. In addition, a common criticism received by these methods is
that the basis function coefficients tend to be overly penalized due to
the sparsity assumption directly applied on the basis function
coefficients.\citep{Scheipl2013} It would be of special interest to
develop a fast yet flexible generalized additive model.

To address these challenges, we propose a novel Bayesian hierarchical
generalized additive model (BHAM) for high dimensional data analysis.
Specifically, we impose a new bi-part spike-and-slab spline prior on the
smooth functions to perform bilevel selection, where the linear and
nonlinear spaces of smooth functions can be selected separately. The
prior setup encourages a flexible solution, rending one of three
possibilities: no effect, only linear effect, or nonlinear effect of a
predictor. Additionally, two fast computing optimization-based
algorithms are developed for high and ultra-high dimension settings
respectively. Meanwhile, to avoid the over-penalization of the basis
function coefficients, smoothness penalties are incorporated in the
model via re-parameterization. Smoothness penalties are commonly
implemented in GAMs through smoothing and penalized regression splines.
They are quadratic norms of the coefficients and allow locally adaptive
penalties on each smoothing function. For the most common cases of cubic
splines, the smoothness penalty conditioning on a smoothness penalty
parameter \(\lambda_j\) is a function of the integration of the second
derivative of the spline function, expressed mathematically as
\begin{equation}\label{eq:smoothpen}
  \text{pen}\left[f_j(x)\right] = \lambda_j \int f^{"}_j(x)dx = \lambda_j \boldsymbol{\beta}_j^T \boldsymbol{S}_j \boldsymbol{\beta}_j ,
\end{equation} where \(\boldsymbol{S}_j\) is a known smoothness penalty
matrix, and \(\boldsymbol{\beta}_j\) is the basis function coefficients.
Smoothness penalties are also previously used in the spike-and-slab GAM
\citep{Scheipl2012} and the sparsity-smoothness penalty
\citep{Meier2009}. Moreover, incorporating the smoothness penalty allows
the separation of the linear space of the smooth function, also know as
the null space, from the nonlinear space. \citet{Marra2011} introduced a
re-parameterization of the basis function matrix based on the
decomposition of the smoothness penalty matrix \(\boldsymbol{S}\). The
re-parameterization allows different penalties imposed on the null and
nonlinear spaces separately, and further allows the sparsity in linear
space and smoothness in nonlinear effect interpolation. \footnote{TODO:
  a sentence that it decompose the prior to independent priors, provide
  obvious computational convenience.} The computational convenience
improved the scalability of the model.

The proposed BHAM uses a spike-and-slab prior for function selections,
but it substantially differs from previous spike-and-slab based GAMs,
i.e.~the spike-and-slab GAM \citep{Scheipl2012} and the SB GAM
\citep{Bai2021} in many ways. First of all, the proposed spike-and-slab
spline prior is a spike-and-slab lasso type prior using independent
mixture double exponential distribution, compared to spike-and-slab GAM.
Spike-and-slab lasso priors provide computational convenience during
model fitting through optimization algorithms, instead of computational
intensive sampling algorithms. They make fitting high- and
ultrahigh-dimensional models feasible without sacrificing performance in
prediction and variable selection. Secondly, the optimization algorithms
introduced in this article include EM-Coordinate Descent algorithm and
EM-Iterative weighted least square algorithm for different utilities.
The applications of the two aforementioned algorithm are fruitful in
Bayesian hierarchical models in high-dimensional data analysis, see
\footnote{TODO: add citations}. Even though SB-GAM with a group
spike-and-slab lasso prior also uses EM-CD for model fitting, the
proposed EM-IWLS provides uncertainty measures that could be further
used for confidence bound calculation and hypothesis testing. EM-IWLS
complements the lack of inference potential from EM-CD in a median and
high data setting. Furthermore, the proposed model addresses the
incapability of bilevel selection in SB-GAM. Lastly, the R
implementation of the Bayesian hierarchical additive model is available
publicly via BHAM at \url{https://github.com/boyiguo1/BHAM}, make
transnational science more accessible.

The rest of the paper are arranged as the followings: in Chapter 2 we
demonstrate the Bayesian hierarchical generalized additive model, where
we provide the setup of the proposed model and describe the two
fast-fitting algorithms. In Chapter 3, we compare the proposed via monte
carlo simulation. .Application to real-world datasets demonstrates the
usefulness of the proposed methodologies in Chpater 4. Conclusion and
discussions are made in Chapter 5.

@@

Following the GLM notation introduced in Section \label{introduction},
we have a generalized additive model and its matrix form,
\begin{equation}\label{eq:gam}
E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}f_j(x_j)) = g^{-1}(a + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j),
\end{equation} with the smoothing function \(f_j(x_j)\) of the variable
\(x_j, j = 1, \dots, p.\) The basis function matrix, i.e.~the design
matrix derived from the smoothing function \(f_j(x_j)\), is denoted
\(\boldsymbol{X}_j\) for the variable \(x_j\). The dimension of the
design matrix depends on the choice of the smoothing function. For
example, a cubic spline function with \(k\) knots produces a \(k+2\)
elements column vector. Its corresponding smoothness penalty matrix is
denoted as \(\boldsymbol{S}_j\). \(\boldsymbol{\beta}_j\) denotes the
basis function coefficients for the \(j\)th variable, and the matrix
transposing operation is denoted with a superscript \(^T\). To note, the
proposed model can include parametric forms of variables in the model,
treating the parametric function as a special case of the smoothing
function, e.g.~\(f_j(x_j) = x_j\) with the smoothness penalty matrix
defined as \(\boldsymbol{S}_j = \begin{bmatrix}0\end{bmatrix}\).

To encourage proper smoothness of the smoothing functions, we adopt the
idea of smoothing penalties from smoothing spline models. The basic idea
is to set up a smoothing penalty, depending on the smoothness penalty
matrix via Equation \ref{eq:smoothpen}, in the posterior density
function, or least squared and likelihood function in the penalized
framework. However, the direct integration of smoothness penalties into
sparsity penalties is not obvious. Given the smoothness penalty matrix
is most often symmetric and positive semi-definite \footnote{ we limit
  the discussion to univariate smoothing function. However, the penalty
  matrix could possibly contain multiple matrices, e.g.~tensor product
  smoother.}, it can be integrated into the design matrix
\(\boldsymbol{X}\) via a re-parameterization. The re-parameterization
follows \citet{Marra2011}, in which eigen-decomposed the penalty matrix
\(\boldsymbol{S} = \boldsymbol{U} \boldsymbol{D} \boldsymbol{U}^T\),
where the matrix \(\boldsymbol{D}\) is diagonal with the eigenvalues
arranged in the ascending order. To note, \(\boldsymbol{D}\) can contain
\(0\) elements on the diagonal, associated with the null space of the
smooth function. We further write the orthonormal matrix
\(\boldsymbol{U}\), containing the eigenvectors as column, as an (row)
augmentation of two sub-matrices
\(\boldsymbol{U} \equiv \begin{bmatrix}\boldsymbol{U}^0 : \boldsymbol{U}^{\text{pen}}\end{bmatrix}\),
where \(\boldsymbol{U}^{\text{pen}}\) contains the eigenvectors (as
columns) for the non-zero eigenvalues, i.e.~the penalized space, and
\(\boldsymbol{U}^0\) contains the eigenvectors for the null space. We
multiply the basis function matrix \(\boldsymbol{X}\) with the
orthonormal matrix \(\boldsymbol{U}\) for the new design matrix
\(\boldsymbol{X}^\ast = \boldsymbol{X} \boldsymbol{U} \equiv \begin{bmatrix} \boldsymbol{X}^0 : \boldsymbol{X}^{\text{pen}} \end{bmatrix}\).
An additional scaling step is imposed on \(\boldsymbol{X}^{\text{pen}}\)
by the non-zero eigenvalues of \(\boldsymbol{D}\) such that the new
basis function matrix \(\boldsymbol{X}^\ast\) is scale-invariant
\footnote{ I know this work for tensor spline (see \citet{Wood2006}),
  but I am not sure if it can be generalized to all smoothing functions}.
For notation convenience, we denote
\(\boldsymbol{X}_j \equiv \begin{bmatrix} \boldsymbol{X}_j^0 : \boldsymbol{X}_j^{\text{pen}} \end{bmatrix}\)
as the after re-parameterized basis function matrix for the \(j\)th
variable hereafter. Similarly, the spline function can be expressed in
the matrix form \[
f_j(x_j) = f_j^0(x_j) + f_j^\text{pen}(x_j) = \boldsymbol{\beta_j^0}^T \boldsymbol{X}_j^0 + \boldsymbol{\beta_j^\text{pen}}^T \boldsymbol{X}_j^\text{pen},
\] and the generalized additive model in Equation \ref{eq:gam} be
\begin{equation}\label{eq:gam-repa}
E(Y|\boldsymbol{x}) =  g^{-1}(a + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j) = g^{-1}\left[a + \sum\limits^p_{j=1} ({\beta_j^0}^T \boldsymbol{X}_j^0 + {\beta_j^\text{pen}}^T \boldsymbol{X}_j^\text{pen})\right].
\end{equation}

@@

The family of spike-and-slab (SS) priors regression models dominates
Bayesian high-dimensional analysis for its utility in outcome prediction
and variable selection. We defer to \citet{Bai2021Review} for an
in-depth introduction to spike-and-slab priors. To summary,
spike-and-slab priors are a family of mixture distributions that
comprises a skinny spike density \(f_{spike}(\cdot)\) for weak signals
and a flat slab density \(f_{slab}(\cdot)\) for large signals,
mathematically \[
 \beta|\gamma \sim (1-\gamma)f_{spike}(\beta) + \gamma f_{slab}(\beta).
\] The most distinct feature of the SS priors is that it is conditioned
on a latent binary variable \(\gamma\) that indicates whether the
variable \(x\) is included in the model. There are various
spike-and-slab priors depending on the choice for the spike density
\(f_{spike}(\cdot)\)\footnote{TODO: should I change to notation of \(f\)
  to not confound with the smooth functions?} and the slab density
\(f_{slab}(\cdot)\), for example \citet{George1993}; \citet{George1997};
\citet{Chipman1996} for grouped variables; \citet{Brown1998} for
multivariate outcomes; \citet{Ishwaran2005}; \citet{Clyde2004} and
reference therein.

The major criticism of early spike-and-slab priors receives is being
computationally prohibitive. \citep{Bai2021Review} Since then, many
studies focus on alleviating the computational burden that sampling
algorithms bear, which include EMVS \citep{Rockova2014a}, \footnote{TODO:
  add more citation}. Particularly, the development of Spike-and-slab
Lasso \citep{Rockova2018b, Rockova2018} substantially improves the
scalability of SS models, setting up the theoretical foundation for
generalized models in -omics data analysis \footnote{TODO: add Dr.~Yi's
  papers}. The spike-and-slab lasso prior is composed of two double
exponential distributions with mean 0 and different dispersion
parameters, \(0 < s_0 < s_1\), mathematically,
\begin{equation} \label{eq:ssl}
\beta | \gamma \sim (1-\gamma)DE(0, s_0) + \gamma DE(0, s_1), 0 < s_0 < s_1.
\end{equation} The SSL also mitigated the problem of EMVS where the weak
signals are not shrink to zero, favored in high-dimensional data
analysis. We noticed that \citet{Bai2021} introduces a spike-and-slab
group prior in the GAM framework, where the densities of the spike and
slab components take the group lasso density \citep{Xu2015}.
\citet{Bai2021} takes an ``all-in-all-out'' strategy for function
selection.

@@

where \(\gamma^0_{j}\in\{0,1\}\) and
\(\gamma^\text{pen}_{j}\in \{0,1\}^{K_j}\) are latent indicator
variables, indicating if the model includes the linear effect and the
non-linear effect of the \(j\)th variable respectively. ; \(s_0\) and
\(s_1\) are scale parameters, assuming \(0 < s_0 < s_1\) and given.
These scale parameters \(s_0\) and \(s_1\) can be treated as tuning
parameters and optimized via cross-validation. A discussion of how to
choose the scale parameters comes in Section
\ref{selecting-optimal-scale-values}. To note, this prior differs from
previous group spike-and-slab lasso priors{[}TODO: add Dr.~Yi's
citation{]}, where the \(\boldsymbol{\beta}^0_j\) and
\(\boldsymbol{\beta}^\text{pen}_j\) shares the same indicator variables
\(\gamma_j^0\), \(\gamma_j^\text{pen}\) respectively. It is possible to
relax the assumption such that each coefficient \(\beta\) to have its
own latent indicator \(\gamma\), but at the cost ofcomplicating the
bi-level function selection. Conversely, it is possible to add a more
restrictive assumption on the priors, assuming that one indicator
variable decides the inclusion of both the linear effect and non-linear
effect, i.e.~\(\gamma_j = \gamma^0_j = \gamma^\text{pen}_j\). This
converges\footnote{Is ``becomes'' better?} to the SB-GAM
\citep{Bai2021}.

The reparameterization introduced in Section
\ref{bayesian-hierarchical-additive-models} grants the validity of the
proposed prior. First of all, the smooth function bases are linear
dependent to each other based on the definitions, and ask for extra
attention. The eigeondecomposition remedies the problem. Hence our prior
can be set to be conditionally independent. Secondly, the eigenvalue
scaling provides a panacea to allow unified scale parameters for all
bases of all smoothing functions of variables, where scaling is mandate.

The rest of the hierchical prior follows the traditional spike-and-slab
lasso piror: we set up the hyper-prior of \(\boldsymbol{\gamma}\), to
allow local adaption of the shrinkage, using a bernoulli distribution.
The two indicators of the \(j\)th predictor, \(\gamma^{0}_j\) and
\(\gamma^\text{pen}_j\), shares the same probability parameter, \[
\begin{aligned}
\gamma_{j}^{0} | \theta_j &\sim Bin(\gamma^{0}_{j}|1, \theta_j)\\
\gamma_{j}^\text{pen}| \theta_j &\sim Bin(\gamma^\text{pen}_{j}|1, \theta_j).
\end{aligned}
\]

>>>>

\label{sec:intro}

Using additive models to interpolate nonlinear effects has a long
history. Early work of additive models can date back to 1979 when
Cleveland \citep{Cleveland1979} proposed local smoothing for bivariate
analysis. Since then, many mathematical and statistical models have been
proposed. \citep{Wood2020} With the recent advancement of computational
technology, many machine learning algorithms that are computationally
intensive start to gain much popularity in prognostic modeling and
functional selection, especially when complex signals, for example
nonlinear effects, are assumed. Such models generally have improved
prediction accuracy over classic statistical models. However, these
models have a fatal flaw, the lack of transparency and interpretability,
and hence are referred to as ``black box'' models \citep{Breiman2001}.
Given that the primary objective of biomedical studies is to understand
the etiology of diseases, ``Black box'' models are less helpful as they
can hardly explain how the change of biomarkers is associated with the
disease.

Semi-parametric regression models are a more appropriate alternative to
the ``black-box'' models, thanks to their balancing between model
flexibility and interpretability. Among those, generalized additive
models (GAMs), proposed in the seminal work of Hastie and Tibshirani
\citep{Hastie1987}, grew to be one of the most popular modeling tools.
In a GAM, the response variable, \(Y\), which is assumed to follow some
exponential family distribution \(EF(\mu, \phi)\) with mean \(\mu\) and
dispersion \(\phi\), can be expressed as the summation of smooth
functions, \(f_j, j = 1, \dots, p\), of a given \(p\)-dimensional vector
of covariates \(\boldsymbol{x}\), written as \[
 E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}f_j(x_j)),
\] where \(g^{-1}\) is the inverse function of a smooth monotonic link
function. The smooth functions can take many forms and are estimated
using a pseudo-weighted version of the backfitting algorithm
\citep{Breiman1985}. Nevertheless, the classic GAMs cannot fulfill the
increasing analytic demands for high-dimensional data analysis in
biomedical studies.

Many modern biomedical research, e.g.~sequencing data analysis, electric
health record data analysis, require special treatment of
high-dimensionality, commonly known as \(p >> n\) problem. There is
extensive literature on high-dimensional linear models via penalized
models or Bayesian hierarchical models, see Mallick and Yi
\citep{Mallick2013} for review. These models assume sparsity among the
predictors, i.e.~most predictors have small or no effects on the
outcome. Further development focuses on addressing the grouping
structure among the predictors, where AMs and GAMs can be considered
special cases. Various penalties are imposed on the basis function
coefficients: Ravikumar et al. \citep{Ravikumar2009} extended the
grouped lasso \citep{Yuan2006} to additive models; Huang et al.
\citep{Huang2010} further developed adaptive grouped lasso for additive
models; Wang et al. \citep{Wang2007} and Xue \citep{Xue2009}
respectively applied grouped SCAD penalty \citep{Fan2001} to the
additive models. Recent Bayesian hierarchical models are also used in
the context of high-dimensional additive models. Various group
spike-and-slab priors combining with computationally intensive Markov
chain Monte Carlo (MCMC) algorithms \citep{Xu2015, Yang2020} are
proposed, where the application on AMs are by-products. Scheipl et al.
\citep{Scheipl2012} proposed an ad hoc spike-and-slab structure prior,
normal-mixture-of-inverse gamma, specifically for semi-parametric
regression using MCMC. The normal-mixture-of-inverse gamma prior is
mixture of t-distribution with an additional coefficients mixing vector,
where the vector follows a mixture of two Gaussian distribution. Bai et
al. \citep{Bai2020} was the first to apply group spike-and-slab lasso
prior to AMs using a fast optimization algorithm, and further
generalized the framework to GAMs\citep{Bai2021}. Regardless of the nice
statistical properties, a practical and pivotal question, if non-linear
effects exist, yet to be addressed by these methods. Besides Scheipl et
al. \citep{Scheipl2012}, most proposed methods adapt an `all-in-all-out'
strategy, i.e.~either including or excluding the variable completely,
rendering no space for within group selection. In addition, a common
criticism received by these methods is that the basis function
coefficients tend to be overly penalized due to the sparsity assumption
directly applied on the basis function coefficients.\citep{Scheipl2013}
It would be of special interest to develop a fast yet flexible
generalized additive model.

To address these challenges, we propose a novel Bayesian hierarchical
generalized additive model (BHAM) for high dimensional data analysis.
Specifically, we impose a new bi-part spike-and-slab spline prior on the
smooth functions to perform bilevel selection, where the linear and
nonlinear spaces of smooth functions can be selected separately. The
prior setup encourages a flexible solution, rending one of three
possibilities: no effect, only linear effect, or nonlinear effect of a
predictor. Additionally, two fast computing optimization-based
algorithms are developed for high and
ultrahigh-dimensional\citep{Fan2009} settings respectively. Meanwhile,
to avoid the over-penalization of the basis function coefficients,
smoothness penalties are incorporated in the model via
re-parameterization. Smoothness penalties are commonly implemented in
GAMs through smoothing and penalized regression splines. They are
quadratic norms of the coefficients and allow locally adaptive penalties
on each smoothing function. For the most common cases of cubic splines,
the smoothness penalty conditioning on a smoothness penalty parameter
\(\lambda_j\) is a function of the integration of the second derivative
of the spline function, expressed mathematically as
\begin{equation}\label{eq:smoothpen}
  \text{pen}\left[f_j(x)\right] = \lambda_j \int f^{"}_j(x)dx = \lambda_j \boldsymbol{\beta}_j^T \boldsymbol{S}_j \boldsymbol{\beta}_j ,
\end{equation} where \(\boldsymbol{S}_j\) is a known smoothness penalty
matrix, and \(\boldsymbol{\beta}_j\) is the basis function coefficients.
Smoothness penalties are also previously used in the spike-and-slab GAM
\citep{Scheipl2012} and the sparsity-smoothness penalty
\citep{Meier2009}. Moreover, incorporating the smoothness penalty allows
the separation of the linear space of the smooth function, also know as
the null space, from the nonlinear space. Marra and Wood
\citep{Marra2011} introduced a re-parameterization of the basis function
matrix based on the decomposition of the smoothness penalty matrix
\(\boldsymbol{S}\). The re-parameterization allows different penalties
imposed on the null and nonlinear spaces separately, and further allows
the sparsity in linear space and smoothness in nonlinear effect
interpolation. In addition, the priors of the linear and nonlinear
effects are independent, creating computational convenience and further
improving the scalability of the model.

The proposed BHAM uses a spike-and-slab prior for function selections,
but it substantially differs from previous spike-and-slab based GAMs,
i.e.~the spike-and-slab GAM \citep{Scheipl2012} and the SB GAM
\citep{Bai2021} in many ways. First of all, the proposed spike-and-slab
spline prior is a spike-and-slab lasso type prior using independent
mixture double exponential distribution, compared to spike-and-slab GAM.
Spike-and-slab lasso priors provide computational convenience during
model fitting through optimization algorithms versus computational
intensive sampling algorithms. They make fitting high- and
ultrahigh-dimensional models more feasible without sacrificing
performance in prediction and variable selection. Secondly, the
optimization algorithms introduced in this article include EM-Coordinate
Descent algorithm and EM-Iterative weighted least square algorithm for
different utilities. The applications of the two aforementioned
algorithm are fruitful in Bayesian hierarchical models in
high-dimensional data analysis.
\citep{Yi2012, Rockova2014a, Rockova2018} Even though SB-GAM with a
group spike-and-slab lasso prior also uses EM-CD for model fitting, the
proposed EM-IWLS provides uncertainty measures that could be further
used for confidence bound calculation and hypothesis testing. EM-IWLS
complements the lack of inference potential from EM-CD in a median and
high dimensional setting. Furthermore, the proposed model addresses the
incapability of bilevel selection in SB-GAM. Lastly, the R
implementation of the Bayesian hierarchical additive model is available
publicly via BHAM at \url{https://github.com/boyiguo1/BHAM}, make
transnational science more accessible.

The rest of the paper are arranged as the followings: in Chapter 2 we
demonstrate the Bayesian hierarchical generalized additive model, where
we provide the setup of the proposed model and describe the two
fast-fitting algorithms. In Chapter 3, we compare the proposed via monte
carlo simulation. .Application to real-world datasets demonstrates the
usefulness of the proposed methodologies in Chpater 4. Conclusion and
discussions are made in Chapter 5. \footnote{TODO: rewrite this
  sentence.}

@@

\label{sec:BHAM}

Following the GLM notation introduced in Section \ref{sec:intro}, we
have a generalized additive model and its matrix form,
\begin{equation}\label{eq:gam}
E(Y|\boldsymbol{x}) = g^{-1}(a + \sum\limits^p_{j=1}f_j(x_j)) = g^{-1}(a + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j),
\end{equation} with the smoothing function \(f_j(x_j)\) of the variable
\(x_j, j = 1, \dots, p.\) The basis function matrix, i.e.~the design
matrix derived from the smoothing function \(f_j(x_j)\), is denoted
\(\boldsymbol{X}_j\) for the variable \(x_j\). The dimension of the
design matrix depends on the choice of the smoothing function. For
example, a cubic spline function with \(k\) knots produces a \(k+2\)
elements column vector. Its corresponding smoothness penalty matrix is
denoted as \(\boldsymbol{S}_j\). \(\boldsymbol{\beta}_j\) denotes the
basis function coefficients for the \(j\)th variable, and the matrix
transposing operation is denoted with a superscript \(^T\). To note, the
proposed model can include parametric forms of variables in the model,
treating the parametric function as a special case of the smoothing
function, e.g.~\(f_j(x_j) = x_j\) with the smoothness penalty matrix
defined as \(\boldsymbol{S}_j = \begin{bmatrix}0\end{bmatrix}\).

To encourage proper smoothness of the smoothing functions, we adopt the
idea of smoothing penalties from smoothing spline models. The basic idea
is to set up a smoothing penalty, depending on the smoothness penalty
matrix via Equation \ref{eq:smoothpen}, in the posterior density
function, or least squared and likelihood function in the penalized
framework. However, the direct integration of smoothness penalties into
sparsity penalties is not obvious. Given the smoothness penalty matrix
is most often symmetric and positive semi-definite \footnote{ we limit
  the discussion to univariate smoothing function. However, the penalty
  matrix could possibly contain multiple matrices, e.g.~tensor product
  smoother.}, it can be integrated into the design matrix
\(\boldsymbol{X}\) via a re-parameterization. The re-parameterization
follows Marra and Wood \citep{Marra2011}, in which eigen-decomposed the
penalty matrix
\(\boldsymbol{S} = \boldsymbol{U} \boldsymbol{D} \boldsymbol{U}^T\),
where the matrix \(\boldsymbol{D}\) is diagonal with the eigenvalues
arranged in the ascending order. To note, \(\boldsymbol{D}\) can contain
\(0\) elements on the diagonal, associated with the null space of the
smooth function. We further write the orthonormal matrix
\(\boldsymbol{U}\), containing the eigenvectors as column, as an (row)
augmentation of two sub-matrices
\(\boldsymbol{U} \equiv \begin{bmatrix}\boldsymbol{U}^0 : \boldsymbol{U}^{\text{pen}}\end{bmatrix}\),
where \(\boldsymbol{U}^{\text{pen}}\) contains the eigenvectors (as
columns) for the non-zero eigenvalues, i.e.~the penalized space, and
\(\boldsymbol{U}^0\) contains the eigenvectors for the null space. We
multiply the basis function matrix \(\boldsymbol{X}\) with the
orthonormal matrix \(\boldsymbol{U}\) for the new design matrix
\(\boldsymbol{X}^\ast = \boldsymbol{X} \boldsymbol{U} \equiv \begin{bmatrix} \boldsymbol{X}^0 : \boldsymbol{X}^{\text{pen}} \end{bmatrix}\).
An additional scaling step is imposed on \(\boldsymbol{X}^{\text{pen}}\)
by the non-zero eigenvalues of \(\boldsymbol{D}\) such that the new
basis function matrix \(\boldsymbol{X}^\ast\) is scale-invariant
\footnote{ I know this work for tensor spline (see \citep{Wood2006}),
  but I am not sure if it can be generalized to all smoothing functions}.
For notation convenience, we denote
\(\boldsymbol{X}_j \equiv \begin{bmatrix} \boldsymbol{X}_j^0 : \boldsymbol{X}_j^{\text{pen}} \end{bmatrix}\)
as the after re-parameterized basis function matrix for the \(j\)th
variable hereafter. Similarly, the spline function can be expressed in
the matrix form \[
f_j(x_j) = f_j^0(x_j) + f_j^\text{pen}(x_j) = \boldsymbol{\beta_j^0}^T \boldsymbol{X}_j^0 + \boldsymbol{\beta_j^\text{pen}}^T \boldsymbol{X}_j^\text{pen},
\] and the generalized additive model in Equation \ref{eq:gam} be
\begin{equation}\label{eq:gam-repa}
E(Y|\boldsymbol{x}) =  g^{-1}(a + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j) = g^{-1}\left[a + \sum\limits^p_{j=1} ({\beta_j^0}^T \boldsymbol{X}_j^0 + {\beta_j^\text{pen}}^T \boldsymbol{X}_j^\text{pen})\right].
\end{equation}

@@

The family of spike-and-slab (SS) priors regression models dominates
Bayesian high-dimensional analysis for its utility in outcome prediction
and variable selection. We defer to Bai et al. \citep{Bai2021Review} for
an in-depth introduction to spike-and-slab priors. To summary,
spike-and-slab priors are a family of mixture distributions that
comprises a skinny spike density \(f_{spike}(\cdot)\) for weak signals
and a flat slab density \(f_{slab}(\cdot)\) for large signals,
mathematically \[
 \beta|\gamma \sim (1-\gamma)f_{spike}(\beta) + \gamma f_{slab}(\beta).
\] The most distinct feature of the SS priors is that it is conditioned
on a latent binary variable \(\gamma\) that indicates whether the
variable \(x\) is included in the model. There are various
spike-and-slab priors depending on the choice for the spike density
\(f_{spike}(\cdot)\)\footnote{TODO: should I change to notation of \(f\)
  to not confound with the smooth functions?} and the slab density
\(f_{slab}(\cdot)\), see George and McCulloch
\citep{George1993, George1997}; Chipman \citep{Chipman1996} for grouped
variables; Brown et al. \citep{Brown1998} for multivariate outcomes;
Ishwaran and Rao \citep{Ishwaran2005}; Clyde and George
\citep{Clyde2004} and reference therein.

The major criticism of early spike-and-slab priors receives is being
computationally prohibitive. \citep{Bai2021Review} Since then, many
studies focus on alleviating the computational burden that sampling
algorithms bear, which include EMVS \citep{Rockova2014a} and
Spike-and-slab Lasso \citep{Rockova2018b, Rockova2018}. Particularly,
the development of Spike-and-slab Lasso substantially improves the
scalability of SS models, setting up the theoretical foundation for
generalized models in -omics data analysis
\citep{Tang2017a, Tang2017, Tang2018, Tang2019}. The spike-and-slab
lasso prior is composed of two double exponential distributions with
mean 0 and different dispersion parameters, \(0 < s_0 < s_1\),
mathematically, \begin{equation} \label{eq:ssl}
\beta | \gamma \sim (1-\gamma)DE(0, s_0) + \gamma DE(0, s_1), 0 < s_0 < s_1.
\end{equation} The SSL also mitigated the problem of EMVS where the weak
signals are not shrink to zero, favored in high-dimensional data
analysis. We noticed that Bai \citep{Bai2021} introduces a
spike-and-slab group prior in the GAM framework, where the densities of
the spike and slab components take the group lasso density
\citep{Xu2015}. Bai \citep{Bai2021} takes an ``all-in-all-out'' strategy
for function selection.

@@

where \(\gamma^0_{j}\in\{0,1\}\) and
\(\gamma^\text{pen}_{j}\in \{0,1\}^{K_j}\) are latent indicator
variables, indicating if the model includes the linear effect and the
non-linear effect of the \(j\)th variable respectively. ; \(s_0\) and
\(s_1\) are scale parameters, assuming \(0 < s_0 < s_1\) and given.
These scale parameters \(s_0\) and \(s_1\) can be treated as tuning
parameters and optimized via cross-validation. A discussion of how to
choose the scale parameters comes in Section \ref{sec:tune}. To note,
this prior differs from previous group spike-and-slab lasso priors
\citep{Tang2018, Tang2019}, where the \(\boldsymbol{\beta}^0_j\) and
\(\boldsymbol{\beta}^\text{pen}_j\) shares the same indicator variables
\(\gamma_j^0\), \(\gamma_j^\text{pen}\) respectively. It is possible to
relax the assumption such that each coefficient \(\beta\) to have its
own latent indicator \(\gamma\), but at the cost ofcomplicating the
bi-level function selection. Conversely, it is possible to add a more
restrictive assumption on the priors, assuming that one indicator
variable decides the inclusion of both the linear effect and non-linear
effect, i.e.~\(\gamma_j = \gamma^0_j = \gamma^\text{pen}_j\). This
converges\footnote{Is ``becomes'' better?} to the SB-GAM
\citep{Bai2021}.

The reparameterization introduced in Section \ref{sec:BHAM} grants the
validity of the proposed prior. First of all, the smooth function bases
are linear dependent to each other based on the definitions, and ask for
extra attention. The eigeondecomposition remedies the problem. Hence our
prior can be set to be conditionally independent. Secondly, the
eigenvalue scaling provides a panacea to allow unified scale parameters
for all bases of all smoothing functions of variables, where scaling is
mandate.

The rest of the hierarchical prior follows the traditional
spike-and-slab lasso piror: we set up the hyper-prior of
\(\boldsymbol{\gamma}\), to allow local adaption of the shrinkage, using
a bernoulli distribution. The two indicators of the \(j\)th predictor,
\(\gamma^{0}_j\) and \(\gamma^\text{pen}_j\), shares the same
probability parameter, \[
\begin{aligned}
\gamma_{j}^{0} | \theta_j &\sim Bin(\gamma^{0}_{j}|1, \theta_j)\\
\gamma_{j}^\text{pen}| \theta_j &\sim Bin(\gamma^\text{pen}_{j}|1, \theta_j).
\end{aligned}
\]
